%% ==========================================================================
%% NeuralOMT — 学习驱动的内部布尔分支 (AAAI 2027 匿名投稿草稿)
%%
%% 编译:  pdflatex main ; bibtex main ; pdflatex main ; pdflatex main
%% 依赖:  aaai2027.sty, aaai2027.bst, references.bib (均在本目录)
%%
%% 说明:实验表格已按论文结构写好、结果单元格留空(\blank),待最终跑批填入。
%% ==========================================================================
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2027}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}

% 数学环境(标准包,AAAI 允许)
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\pdfinfo{
/TemplateVersion (2027.1)
}

\setcounter{secnumdepth}{2} % 采用编号章节(形式化论文风格)

% ---- 定理类环境 ----
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

% ---- 常用宏 ----
\newcommand{\blank}{\,\rule[-.1ex]{2.2em}{0.4pt}\,} % 留空的结果单元格
\newcommand{\solver}{\textsc{z3}}
\newcommand{\Imp}{\mathrm{Imp}}
\newcommand{\score}{\mathrm{score}}
\newcommand{\decide}{\textsf{add\_decide}}
\newcommand{\nextsplit}{\textsf{next\_split}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Atoms}{\mathcal{A}}
\newcommand{\Undec}{U}
\newcommand{\defer}{\bot}
\newcommand{\policy}{\pi_\theta}
\newcommand{\confl}{\texttt{conflicts}}
\newcommand{\rlimit}{\texttt{rlimit}}
\DeclareMathOperator*{\meanop}{mean}

% ---- 标题(mixed case) ----
\title{When Does Learned Boolean Branching Help Inside {CDCL(T)}?\\
A Non-Intrusive Neural Mechanism and an Honest Study}

\author{
    Anonymous Submission
}
\affiliations{
    Paper under double-blind review
}

\begin{document}

\maketitle

\begin{abstract}
Modern SMT and Optimization Modulo Theories (OMT) solvers spend much of
their effort in a \textsc{CDCL(T)} search whose efficiency hinges on the
Boolean branching heuristic (typically VSIDS). Whether a \emph{learned}
heuristic can improve on VSIDS \emph{inside a production solver, without
modifying it}, remains largely open---especially beyond pure propositional
SAT. We present a non-intrusive mechanism that injects a learned decision
policy into the internal \textsc{CDCL(T)} decisions of an unmodified
\solver{} through its \textsc{UserPropagator} interface, and we formalize
it as a \emph{decision oracle} over a black-box \textsc{CDCL(T)} procedure.
We prove that the mechanism preserves soundness and completeness, and we
give a measurement protocol that isolates branching quality by attaching a
(possibly inert) propagator to \emph{both} compared arms, so that they
explore the same unpreprocessed search space. On top of the mechanism we
build a heterogeneous-graph neural policy trained by a consequence-based
look-ahead teacher (imitation) and by REINFORCE with a per-instance
baseline. We study the policy on hard random 3-SAT, on SMT(LIA) instances
with rich theory-atom structure, and in the OMT optimization loop. Our
central finding is a \emph{boundary}: a learned quasi-static, full-coverage
decision order improves on VSIDS on pure-Boolean SAT, but does \emph{not}
on theory atoms or in the OMT loop; a systematic root-cause analysis
attributes the gap to VSIDS's conflict-adaptive dynamics rather than to the
teacher signal. We release the instance generators, harnesses, and test
suite. \emph{(Empirical tables are scaffolded; result cells are to be filled
from final runs.)}
\end{abstract}

% ==========================================================================
\section{Introduction}
% ==========================================================================

Satisfiability Modulo Theories (SMT) solvers are the workhorses of
verification, synthesis, and constraint solving, and Optimization Modulo
Theories (OMT) extends them to problems whose solutions must be optimal with
respect to an objective~\citep{barrett2021smt,sebastiani2015pushing,bjorner2015nz}.
At the heart of a modern solver lies a \textsc{CDCL(T)} loop
\citep{nieuwenhuis2006dpllt}: a conflict-driven clause-learning SAT engine
over the Boolean abstraction of the formula, tightly coupled with theory
propagation. The efficiency of that loop is famously sensitive to one
component---the \emph{branching heuristic} that chooses which Boolean atom
to case-split on and with what phase. The de-facto standard, VSIDS
\citep{moskewicz2001chaff}, is a cheap, conflict-adaptive scoring rule that
has resisted improvement for two decades.

Machine learning has repeatedly been proposed to \emph{learn} better
branching decisions. In mixed-integer programming, graph neural networks
imitate strong branching with striking success
\citep{gasse2019exact,nair2020solving}; in propositional SAT, neural models
either learn end-to-end solvers \citep{selsam2019neurosat} or learn a
branching policy by reinforcement learning \citep{kurin2020graphqsat}. Yet
two gaps persist. \emph{First}, almost all of this work either builds a
bespoke research solver or patches the source of an existing one, whereas
production SMT/OMT solvers are complex, highly tuned black boxes that
practitioners are reluctant to fork. \emph{Second}, the evidence is
concentrated on a single regime---pure SAT or pure MILP---and says little
about whether learned branching transfers to the \emph{theory} reasoning of
\textsc{CDCL(T)} or to the outer optimization loop of OMT.

This paper addresses both gaps. We observe that a mature solver,
\solver{}~\citep{demoura2008z3}, exposes a \emph{supported, non-intrusive}
seam---the \textsc{UserPropagator} API, whose \decide{}/\nextsplit{}
callbacks let an external policy dictate the solver's next internal
\textsc{CDCL(T)} decision without recompiling the solver. We turn this
engineering seam into a clean formal object: a \emph{decision oracle} over a
black-box \textsc{CDCL(T)} procedure. We prove that any oracle of the right
shape preserves soundness and completeness, and we identify a subtle but
crucial measurement issue: attaching a propagator disables \solver{}'s
preprocessing, so a fair comparison must attach a (possibly inert)
propagator to \emph{both} arms. On top of the mechanism we build a
graph-neural policy trained by a look-ahead teacher and reinforcement
learning, and we run it across the SAT $\to$ SMT(LIA) $\to$ OMT spectrum.

The result is an honest characterization rather than a uniform win. On hard
random 3-SAT the learned policy reduces conflicts relative to VSIDS; on
SMT(LIA) theory atoms and in the OMT loop it does not. A systematic probe
localizes the cause: the \emph{learnability} of the look-ahead signal and
the \emph{utility} of the resulting policy are separate questions---a
learned quasi-static full-coverage order can beat VSIDS on a purely
propositional clause graph, but it loses to VSIDS's conflict-adaptive
dynamics once decisions range over theory atoms, and branching is simply not
the bottleneck inside the OMT linear-search loop.

\paragraph{Contributions.}
\begin{itemize}
\item \textbf{A non-intrusive formal mechanism} (Section~\ref{sec:mechanism}):
we model external control of a black-box \textsc{CDCL(T)} solver's internal
branching as a \emph{decision oracle} $\delta$, realize it via
\solver{}'s \textsc{UserPropagator}, and prove it preserves soundness and
completeness (Proposition~\ref{prop:sc}). We give a measurement protocol
that isolates branching quality (Section~\ref{sec:isolation}) and show how
OMT is handled as a degenerate Generalized-OMT linear search.
\item \textbf{A neural instantiation} (Section~\ref{sec:policy}): a
heterogeneous-graph GNN policy with a consequence-based look-ahead teacher
for imitation cold-start and a REINFORCE objective with a per-instance
baseline, including a theory-atom structural encoding that makes the
imitation signal learnable.
\item \textbf{An honest empirical study} (Sections~\ref{sec:setup}--\ref{sec:results}):
a paired, seed-disjoint protocol across hard 3-SAT, SMT(LIA), and the OMT
loop, with a root-cause analysis separating learnability from solving
utility.
\item \textbf{Reusable artifacts}: instance generators, both evaluation
harnesses, and a passing test suite, released to support replication.
\end{itemize}

% ==========================================================================
\section{Background and Related Work}
% ==========================================================================

\paragraph{SMT, CDCL(T), and OMT.}
Fix a background theory $\mathcal{T}$. A quantifier-free formula $\varphi$
has a \emph{Boolean abstraction} obtained by replacing each theory atom
(e.g.\ a linear-arithmetic comparison $\sum_i c_i x_i \bowtie b$) with a
fresh Boolean variable. \textsc{CDCL(T)} \citep{nieuwenhuis2006dpllt}
searches for a theory-consistent truth assignment to these atoms by
interleaving Boolean unit propagation, theory propagation, and
conflict-driven clause learning. At each \emph{decision point} the solver
picks an unassigned atom and a phase; VSIDS \citep{moskewicz2001chaff}
scores atoms by recent conflict participation, and refinements such as LRB
\citep{liang2016lrb} learn the scoring online. \emph{Look-ahead} solvers
\citep{heule2021lookahead} instead score a literal by how much propagation
it triggers---the intuition our teacher signal formalizes. OMT
\citep{sebastiani2015pushing,bjorner2015nz,trentin2019omt} augments SMT with
an objective term to be optimized, typically by a linear or binary search
that repeatedly tightens a bound and re-checks satisfiability.

\paragraph{Generalized OMT.}
\citet{gomt2024} give a theory-agnostic abstract calculus for OMT (GOMT)
whose state $\langle \mathcal{I}, \Delta, \tau\rangle$ is manipulated by
three derivation rules (\textsc{F-Split}, \textsc{F-Sat}, \textsc{F-Close})
around a \textsc{Solve} oracle. A \emph{linear search} strategy in that
calculus never splits and applies \textsc{F-Sat} until saturation. Our OMT
harness is exactly this degenerate strategy: a \solver{} \texttt{Solver}
loop that repeatedly checks $\varphi \wedge (\text{bound})$ and tightens the
bound. This matters because \solver{}'s \texttt{Optimize} engine does
\emph{not} accept a propagator, so learned branching can only be attached to
the \texttt{Solver} checks inside the loop---placing our mechanism squarely
at the \textsc{Solve} oracle of the GOMT calculus rather than at its outer
branch ordering.

\paragraph{Learning to branch.}
The closest lines of work are neural branching for MILP
\citep{gasse2019exact,nair2020solving} and for SAT
\citep{selsam2019neurosat,kurin2020graphqsat}. We differ in three ways: (i)
we target the \emph{theory}-level search of \textsc{CDCL(T)} and the OMT
loop, not only propositional SAT or MILP; (ii) we inject the policy into an
\emph{unmodified, production} solver through a supported API, rather than
forking the solver; and (iii) we report a boundary result with a root-cause
analysis, rather than a single-regime win. Methodologically we build on
relational graph convolution \citep{schlichtkrull2018rgcn}, listwise ranking
\citep{cao2007listnet}, and policy-gradient RL \citep{williams1992reinforce}.

% ==========================================================================
\section{A Non-Intrusive Mechanism for Learned Branching}
\label{sec:mechanism}
% ==========================================================================

\subsection{The Decision-Oracle Abstraction}

Let $S$ be a \textsc{CDCL(T)} decision procedure for $\mathcal{T}$-satisfiability
of a quantifier-free formula $\varphi$ whose Boolean abstraction ranges over
a finite atom set $\Atoms(\varphi)$. We model $S$'s search state as a pair
$(M,\Undec)$, where $M$ is the current partial assignment (trail) over
$\Atoms(\varphi)$ and $\Undec\subseteq\Atoms(\varphi)$ is the set of
currently unassigned atoms.

\begin{definition}[Decision oracle]
\label{def:oracle}
A \emph{decision oracle} is a function $\delta$ that maps a search state
$(M,\Undec)$ either to a \emph{decision literal} $\ell$ over some atom
$a\in\Undec$ (i.e.\ $a$ together with a phase), or to the distinguished value
$\defer$ (``defer''). An oracle is \emph{well-formed} if it never returns a
literal over an atom outside $\Undec$: it only chooses \emph{which
unassigned atom to case-split next, and with what phase}; it adds and removes
no atoms, clauses, or constraints.
\end{definition}

Write $S[\delta]$ for the procedure $S$ run so that, at every decision point,
it decides $\ell$ next if $\delta(M,\Undec)=\ell$, and otherwise
($\delta(M,\Undec)=\defer$) it decides using its native heuristic $h$
(VSIDS). Algorithm~\ref{alg:cdclt} shows the decorated loop; the only change
from stock \textsc{CDCL(T)} is line~\ref{line:oracle}.

\begin{algorithm}[t]
\caption{\textsc{CDCL(T)} with a decision oracle}
\label{alg:cdclt}
\begin{algorithmic}[1]
\REQUIRE formula $\varphi$ over atoms $\Atoms$; native heuristic $h$; oracle $\delta$
\STATE attach propagator $\Rightarrow$ disable preprocessing; $M \gets \emptyset$
\LOOP
  \STATE $(M,\mathit{st}) \gets \textsc{Propagate}(\varphi, M)$ \COMMENT{unit + theory propagation}
  \IF{$\mathit{st} = \textsc{conflict}$}
     \IF{top-level conflict} \RETURN \textsc{unsat} \ENDIF
     \STATE analyze, learn clause, backjump$(M)$; \textbf{continue}
  \ENDIF
  \STATE $\Undec \gets \{a\in\Atoms : a \text{ unassigned in } M\}$
  \IF{$\Undec = \emptyset$} \RETURN \textsc{sat} (model $M$) \ENDIF
  \STATE $\ell \gets \delta(M,\Undec)$ \label{line:oracle} \COMMENT{learned override point}
  \IF{$\ell = \defer$} \STATE $\ell \gets h(M,\Undec)$ \COMMENT{fall back to VSIDS} \ENDIF
  \STATE $M \gets M \cup \{\,\text{decide } \ell\,\}$
\ENDLOOP
\end{algorithmic}
\end{algorithm}

\subsection{Realization via \textsc{UserPropagator}}

\solver{} exposes \texttt{UserPropagateBase} with two hooks that implement
Definition~\ref{def:oracle} without touching solver internals:
\decide{}, invoked at each internal decision point, and \nextsplit{}$(a,
\text{phase})$, by which the callback names the next literal to branch on.
Returning without a call (``$\delta=\defer$'') hands control back to native
VSIDS. Thus $\ell\gets\delta(M,\Undec)$ in Algorithm~\ref{alg:cdclt} is a
single \decide{} callback, and the fallback of line~13 is automatic.

\begin{remark}[Preprocessing side-effect]
\label{rem:preproc}
Attaching \emph{any} propagator makes \solver{} disable its
preprocessing/in-processing and run a pure \textsc{CDCL(T)} search. This is
central to our study: on preprocessed hard instances the conflict count can
collapse to near zero, erasing the branching signal. Both arms must
therefore attach a propagator (Section~\ref{sec:isolation}).
\end{remark}

\subsection{Soundness and Completeness}

\begin{proposition}[Preservation]
\label{prop:sc}
Let $S$ be a sound and complete \textsc{CDCL(T)} procedure with a fair native
heuristic $h$. For every well-formed decision oracle $\delta$
(Definition~\ref{def:oracle}), $S[\delta]$ terminates and reports
$\mathcal{T}$-sat iff $\varphi$ is $\mathcal{T}$-satisfiable, and
$\mathcal{T}$-unsat otherwise. That is, $S[\delta]$ decides the same problem
as $S$.
\end{proposition}

\begin{proof}[Proof sketch]
\textsc{CDCL(T)} is sound and complete under \emph{any} case-split strategy
that (i) branches only on unassigned atoms of the finite set
$\Atoms(\varphi)$ and (ii) is subject to the usual termination controls
(conflict-driven learning with backjumping, plus a fair restart/decision
policy so that no atom is starved). A well-formed $\delta$ satisfies (i) by
construction, since it returns a literal only over an atom in $\Undec$. For
(ii): whenever $\delta=\defer$ the fair native heuristic $h$ is used, and
even when $\delta\neq\defer$ its choice comes from $\Undec$, so on every
branch $\Undec$ strictly shrinks toward either a full model or a conflict;
conflict analysis, clause learning, and backjumping are untouched. Decision
\emph{order and phase} affect which conflicts are encountered first, hence
runtime, but not the \emph{existence} of a satisfying assignment or of a
refutation: a returned model still satisfies $\varphi$ and is
theory-consistent, and an \textsc{unsat} answer is still certified by a
resolution/theory-lemma refutation. Disabling preprocessing
(Remark~\ref{rem:preproc}) changes the search but preserves
sat-equivalence. Hence $S[\delta]$ and $S$ decide the same problem.
\end{proof}

Proposition~\ref{prop:sc} is what makes a \emph{learned} oracle safe: no
matter how badly $\policy$ ranks atoms, correctness is never at risk, and
$\policy=\defer$ recovers the native solver exactly.

\subsection{Isolating Branching Quality}
\label{sec:isolation}

We instrument a single satisfiability check with three quantities returned by
the harness \texttt{solve\_sat\_with\_decider}:
\begin{align*}
\confl &= \text{\solver{} statistic ``conflicts''},\\
\rlimit &= \text{\solver{} statistic ``rlimit count''},\\
\texttt{decisions} &= \#\{\text{steps with } \delta\neq\defer\}.
\end{align*}
Here $\confl$ is the direct proxy for branching quality (fewer is better),
$\rlimit$ is an honest total-work counter that includes theory-solving cost,
and $\texttt{decisions}$ counts only the steps at which the oracle actually
overrode VSIDS.

\begin{definition}[Paired measurement]
\label{def:pairing}
Both arms run $S[\delta]$ on the \emph{same} instance under an attached
propagator: the \emph{baseline} arm uses $\delta\equiv\defer$ (pure native
VSIDS, but with preprocessing disabled), and the \emph{learned} arm uses
$\delta=\policy$. Because a propagator is attached in both arms, they explore
the same unpreprocessed \textsc{CDCL(T)} search space and differ \emph{only}
in the decision heuristic.
\end{definition}

Under Definition~\ref{def:pairing}, the difference in $\confl$ between the two
arms isolates branching quality: it cannot be an artifact of one arm getting
luckier preprocessing, because neither arm is preprocessed. We regard this
protocol as a methodological contribution: it is easy to accidentally compare
a propagator-decorated learned arm against a \emph{preprocessed} native
baseline and thereby measure preprocessing, not branching.

\subsection{OMT as a Degenerate GOMT Linear Search}
\label{sec:omt}

For an OMT problem $\langle t,\prec,\varphi\rangle$ we run the GOMT linear
search of \citet{gomt2024}: repeatedly call \textsc{Solve} on $\varphi$
conjoined with a $\prec$-tightening bound on $t$, updating the incumbent
until the check is unsatisfiable. Since \solver{}'s \texttt{Optimize} rejects
propagators, the oracle attaches to each \texttt{Solver} check in the loop.
We flag one asymmetry honestly: in this harness only the learned arm attaches
a propagator (the baseline uses stock \texttt{Optimize}), so the OMT
comparison does not enjoy the clean isolation of
Definition~\ref{def:pairing}; we return to its consequences in
Section~\ref{sec:results}.

% ==========================================================================
\section{The Neural Branching Policy}
\label{sec:policy}
% ==========================================================================

The oracle $\policy$ is a graph neural network trained in two stages:
imitation of a look-ahead teacher (cold start), then REINFORCE fine-tuning.

\subsection{Look-Ahead Teacher (Imitation Signal)}
\label{sec:teacher}

For each candidate atom $a$ we score its \emph{marginal propagation strength}
using \solver{}'s \texttt{consequences} query, in the spirit of look-ahead
solvers \citep{heule2021lookahead}. Let $\Imp(L)$ be the set of atom keys
forced to a value under the hypothesis set $L$, and let
$B=\Imp(\varnothing)$ be the atoms entailed by the hard constraints alone
(the baseline). Define
\begin{align*}
p_t(a) &= \bigl|\Imp(\{a\})\setminus B\setminus\{a\}\bigr|,\\
p_f(a) &= \bigl|\Imp(\{\neg a\})\setminus B\setminus\{a\}\bigr|,
\end{align*}
and set the teacher score and phase
\begin{equation}
\begin{aligned}
\score(a) &= \bigl(p_t(a)+1\bigr)\bigl(p_f(a)+1\bigr),\\
\mathrm{phase}(a) &= \mathbb{1}[\,p_t(a)\ge p_f(a)\,].
\end{aligned}
\label{eq:teacher}
\end{equation}
The $+1$ bias keeps atoms that propagate nothing comparable (score $1$, not
$0$); subtracting the baseline $B$ is essential, since box/entailed atoms
otherwise ``imply'' the same set under every hypothesis and the labels
degenerate to uniform. Atoms that are already \textsc{unsat} on one side at
the root are skipped: they are resolved by propagation, not by branching. A
configuration \texttt{max\_atoms} caps how many atoms are scored per state.

\begin{remark}
The teacher counts \emph{propagations}, which in SMT includes \emph{theory}
propagation. On pure SAT unit propagation this is a good branching signal;
whether it induces a good \emph{total order} under \textsc{DPLL(T)} is
precisely the empirical question of Section~\ref{sec:results}.
\end{remark}

\subsection{Heterogeneous Graph Encoding}
\label{sec:graph}

A search state is encoded as a heterogeneous graph with six node types and
seven directed edge types (Table~\ref{tab:nodes}). A theory atom
$\sum_i c_i x_i \bowtie b$ is normalized to one side by decomposing both sides
into linear forms $(\mathbf{c}^L,k^L)$ and $(\mathbf{c}^R,k^R)$ and setting
\begin{equation}
\begin{aligned}
\mathbf{c} &= \mathbf{c}^L-\mathbf{c}^R \ (\text{drop zeros}),\quad
b = k^R-k^L,\\
\texttt{kind} &= \mathrm{map}(\bowtie).
\end{aligned}
\label{eq:linear}
\end{equation}
The atom shares a key with its abstracting Boolean variable, so an
\textsc{atom-abstracted-by} edge links the theory-atom node to the Boolean
node, and \textsc{variable-in-atom} edges are emitted from
$\mathbf{c}$. Pure propositional constants carry no theory-atom or
numeric-variable nodes, so a purely Boolean instance yields a purely Boolean
graph. Optional scalars are encoded as $[x,1]$ if present and
$[\text{default},0]$ if absent; large magnitudes pass through
$\mathrm{slog}(x)=\operatorname{sign}(x)\log(1+|x|)$; tri-state Booleans use a
one-hot $[\textit{none},\textit{true},\textit{false}]$.

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}llp{3.9cm}@{}}
\toprule
Node type & Dim & Salient features \\
\midrule
\textsc{bool\_var}     & 19 & assignment, decision level, candidate flag, VSIDS/LRB/CHB activity, phase saving, polarity counts \\
\textsc{clause}        & 11 & clause-type one-hot, LBD, activity, width, satisfied flag \\
\textsc{theory\_atom}  & 22 & relation one-hot, $\mathrm{slog}(b)$, slack/violation/LP value, basis status, arity \\
\textsc{numeric\_var}  & 17 & integrality, LP value/bounds, fractionality, objective coeff., reduced cost, pseudocost \\
\textsc{objective}     & 10 & sense, incumbent/best-bound/gap, lex priority \\
\textsc{search\_state} & 16 & depth, trail, restart/conflict counts, conflict rate, mode \\
\bottomrule
\end{tabular}
\caption{Heterogeneous node types and feature dimensions. Edges are
directed; message passing follows the directed schema (no reverse edges).}
\label{tab:nodes}
\end{table}

\subsection{R-GCN-Style Encoder}
\label{sec:gnn}

We use a relational graph-convolution encoder \citep{schlichtkrull2018rgcn}
with per-type input projection $h_v^{(0)}=\mathrm{Linear}_{t(v)}(x_v)$ and
$L=3$ layers of hidden width $64$. For node $v$ of type $t(v)$,
\begin{equation}
h_v^{(l+1)}=\mathrm{LN}_{t(v)}\!\Bigl(h_v^{(l)}+\mathrm{ReLU}\bigl(
W^{\mathrm{self}}_{t(v)} h_v^{(l)}
+ m_v^{(l)}\bigr)\Bigr),
\label{eq:rgcn}
\end{equation}
\begin{equation}
m_v^{(l)}=\meanop_{\,r,\;u\in\mathcal{N}_r(v)}
W_r\,[\,h_u^{(l)}\,\Vert\,e_{u\to v}\,],
\label{eq:msg}
\end{equation}
where $W_r$ is a per-edge-type linear map whose input concatenates the edge
feature $e_{u\to v}$, messages over all relations $r$ into $v$ are summed and
divided by the in-degree (floored at $1$), and the update applies a
self-transform, a residual connection, and per-type layer normalization.

\subsection{Task Heads and Imitation Loss}
\label{sec:heads}

From the encoded embeddings, a \textsc{BranchingHead} emits a scalar score
$s_a$ per Boolean atom (the main head); a \textsc{PhaseHead} emits a
``branch-true'' logit; an \textsc{IntegerBranchHead} emits a split score and
direction for numeric variables; and \textsc{AuxiliaryHeads} predict
conflict probability, unsat-core membership, objective improvement, and
subtree size. All heads output raw scores; masking/softmax is applied
downstream. The imitation loss is a weighted sum
\begin{equation}
L=L_{\mathrm{branch}}+\lambda_{\mathrm{ph}}L_{\mathrm{ph}}
+\lambda_{\mathrm{int}}(L_{\mathrm{int}}+L_{\mathrm{dir}})
+\lambda_{\mathrm{aux}}L_{\mathrm{aux}},
\label{eq:imit}
\end{equation}
with $\lambda_{\mathrm{ph}}=0.5,\ \lambda_{\mathrm{int}}=1.0,\
\lambda_{\mathrm{aux}}=0.3$. The ranking terms use a ListNet-style
\citep{cao2007listnet} soft cross-entropy between the softmax of the teacher
scores~\eqref{eq:teacher} and the softmax of the head scores,
\begin{equation}
L_{\mathrm{rank}}=-\!\!\sum_{a\in\text{cand}}\!\mathrm{softmax}(\text{tgt})_a\,
\log\mathrm{softmax}(s)_a,
\label{eq:listnet}
\end{equation}
optimized with Adam \citep{kingma2015adam}. Phase/direction use logit BCE;
auxiliary heads use BCE (classification) and MSE (regression).

\begin{remark}[Learnability guard]
Whether $L_{\mathrm{branch}}$ decreases with training is a direct probe of
whether the teacher labels are predictable from the features. On SMT(LIA),
without the theory-atom structural encoding of Section~\ref{sec:graph} the
branch gradient is frozen at the uniform floor $\log N$; adding the
theory-atom features (Eq.~\ref{eq:linear}) restores a monotonically
decreasing branch loss. Learnability, however, is not utility
(Section~\ref{sec:results}).
\end{remark}

\subsection{REINFORCE Fine-Tuning}
\label{sec:rl}

A sampling oracle assigns a distribution over the action set
$[\defer, a_1,\dots,a_m]$ ($m$ current undecided atoms):
\begin{equation}
\policy(\cdot\mid s)=\mathrm{softmax}\bigl([\,\ell_{\defer},s_{a_1},\dots,s_{a_m}\,]\bigr),
\label{eq:rlpolicy}
\end{equation}
where $\ell_{\defer}$ is a shared learnable scalar occupying index $0$;
sampling $\defer$ returns to VSIDS, otherwise the atom is chosen with phase
forced true. The reward is log-compressed, $R=-\log(1+\confl)$ (SAT) or
$-\log(1+\rlimit)$ (OMT), and a per-instance EMA baseline
$b_{\mathrm{key}}\leftarrow0.9\,b_{\mathrm{key}}+0.1\,R$ yields a single
trajectory-level advantage $A=R-b_{\mathrm{key}}$. The policy-gradient loss
\citep{williams1992reinforce} is
\begin{equation}
L_{\mathrm{RL}}=-\frac{1}{n}\sum_{t=1}^{n}\log\policy(a_t\mid s_t)\,\bigl(R-b_{\mathrm{key}}\bigr),
\label{eq:reinforce}
\end{equation}
with no per-step credit assignment and no discounting.

% ==========================================================================
\section{Experimental Setup}
\label{sec:setup}
% ==========================================================================

\paragraph{Instances.}
We use three generators. \emph{Hard random 3-SAT} is drawn at the
phase-transition clause/variable ratio $\approx 4.26$ with $n=70$ variables.
\emph{Pigeonhole (PHP)} gives structured \textsc{unsat} instances. \emph{Hard
SMT(LIA)} is generated by \texttt{generate\_hard\_smt\_lia}: integer
variables $x_i\in[0,\mathrm{ub}]$ under a conjunction of $k$-ary disjunctions
of linear atoms,
\begin{equation}
\bigwedge_{j=1}^{\mathrm{n\_disj}}\ \bigvee_{l=1}^{k}
\Bigl(\textstyle\sum_i c^{(j,l)}_i x_i\ \{\le\ \text{or}\ \ge\}\ b^{(j,l)}\Bigr),
\label{eq:smtgen}
\end{equation}
with coefficients $c\in[-\mathrm{chi},\mathrm{chi}]$ and
$b\in[-\mathrm{ub},\mathrm{ub}\cdot\mathrm{chi}]$, tightened so that Boolean
search becomes the bottleneck (hundreds of conflicts under an attached
propagator). Defaults: $\mathrm{n\_vars}{=}8$, $\mathrm{n\_disj}{=}30$,
$k{=}3$, $\mathrm{ub}{=}6$, $\mathrm{chi}{=}4$.

\paragraph{Protocol.}
For each setting we (i) cold-start by imitation on look-ahead labels
(Eq.~\ref{eq:listnet}), (ii) fine-tune by REINFORCE (Eq.~\ref{eq:reinforce}),
and (iii) evaluate paired on held-out instances under
Definition~\ref{def:pairing}, with \emph{train/test seeds disjoint} to
preclude leakage. We report $\confl$ mean$\pm$std and the paired win-rate
(fraction of instances on which the learned arm has fewer conflicts), and we
include an \emph{untrained} control ($\text{train}{=}0$) to check that any
gain is due to training rather than to the mere presence of a propagator.

\paragraph{Implementation.}
The policy is implemented in PyTorch; the solver is \emph{unmodified}
\solver{}~4.15.4 installed from the \texttt{z3-solver} wheel (no
recompilation). Behavior is governed by configuration dataclasses
(\texttt{LookaheadConfig}, \texttt{TrainConfig}, \texttt{DecideRLConfig})
rather than scattered flags. Hardware and wall-clock budgets: \blank{}
(to be reported).

% ==========================================================================
\section{Results}
\label{sec:results}
% ==========================================================================

Tables~\ref{tab:main}--\ref{tab:ablation} are scaffolded to the final
protocol; result cells (\blank) are to be filled from clean final runs. We
state the \emph{directional} findings, which are robust across our
development runs, and defer magnitudes to the tables.

\begin{table*}[t]
\centering
\small
\begin{tabular}{@{}llcccccc@{}}
\toprule
Setting & Metric & VSIDS & Learned (trained) & Learned (untrained) & Paired win-rate & $\Delta\%$ & $\#$seeds \\
\midrule
Hard 3-SAT ($n{=}70$, $r{\approx}4.26$) & $\confl$ & \blank & \blank & \blank & \blank & \blank & \blank \\
PHP (pigeonhole, \textsc{unsat})        & $\confl$ & \blank & \blank & \blank & \blank & \blank & \blank \\
Hard SMT(LIA) ($8{\times}30$)           & $\confl$ & \blank & \blank & \blank & \blank & \blank & \blank \\
OMT loop (linear search)                & $\rlimit$ & \blank & \blank & \blank & \blank & \blank & \blank \\
\bottomrule
\end{tabular}
\caption{Main paired comparison (baseline VSIDS vs.\ learned oracle). Both
SAT/SMT arms attach a propagator (Definition~\ref{def:pairing}); the OMT row
uses the asymmetric harness of Section~\ref{sec:omt}. Lower is better.}
\label{tab:main}
\end{table*}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Domain & VSIDS & Teacher-oracle & GNN (trained) \\
\midrule
SAT (3-SAT $n{=}70$)  & \blank & \blank & \blank \\
SMT(LIA)              & \blank & \blank & \blank \\
\bottomrule
\end{tabular}
\caption{Root-cause probe: a \emph{teacher-oracle} arm branches by the
look-ahead argmax (Eq.~\ref{eq:teacher}) with no GNN, isolating the teacher
signal from the learned policy. $\confl$; lower is better.}
\label{tab:oracle}
\end{table}

\begin{table}[t]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
Configuration & Branch loss & $\confl$ vs.\ VSIDS \\
\midrule
w/o theory feats.\  & \blank\ (frozen) & \blank \\
w/\ theory feats.\  & \blank\ (drops)  & \blank \\
imitation only      & \blank & \blank \\
imitation $+$ RL    & \blank & \blank \\
\bottomrule
\end{tabular}
\caption{Ablations on SMT(LIA). Without the theory-atom encoding
(Eq.~\ref{eq:linear}) the branch loss is frozen at the uniform floor
$\approx\!\log N$; adding it makes the loss drop (learnability restored). RL
fine-tuning is applied on top of the imitation cold start.}
\label{tab:ablation}
\end{table}

\subsection{Positive: Hard Random 3-SAT}
On hard 3-SAT the learned oracle reduces $\confl$ relative to VSIDS under the
paired protocol (Table~\ref{tab:main}), with the trained policy clearly
ahead of the untrained control. The root-cause probe
(Table~\ref{tab:oracle}) shows the GNN beats \emph{both} VSIDS \emph{and} the
teacher-oracle: the network does not merely imitate the look-ahead teacher
but generalizes a smoother full-coverage order that outperforms it. This is
the source of the SAT gain---not the teacher being strong.

\subsection{Boundary: PHP / \textsc{unsat}}
On pigeonhole \textsc{unsat} instances the learned arm does not help: the
\texttt{consequences} look-ahead teacher provides no labels on top of a
contradictory base (everything is entailed), so imitation has nothing to
cold-start from and RL alone is insufficient. This is an honest limitation of
a propagation-count teacher on \textsc{unsat}.

\subsection{Boundary: SMT(LIA) Theory Atoms}
The theory-atom encoding makes the imitation signal \emph{learnable}
(Table~\ref{tab:ablation}: the branch loss unfreezes from the uniform floor),
yet the learned oracle still loses to VSIDS (Table~\ref{tab:main}), with
large variance across seeds. The probe (Table~\ref{tab:oracle}) localizes the
cause: (i) it is \emph{not} teacher-mismatch---the teacher-oracle is
$\approx$VSIDS on \emph{both} SAT and SMT; (ii) on SAT the GNN generalizes
past the teacher, but on SMT it does not; (iii) the learned policy commits to
a \emph{full-coverage, quasi-static} order (it almost never defers), and such
an order cannot match VSIDS's per-conflict adaptation once decisions range
over theory atoms. In short, a fixed learned order beats VSIDS on a
purely propositional clause graph but is dominated by VSIDS dynamics on
theory atoms.

\subsection{Boundary: The OMT Loop}
In the OMT linear-search loop the learned and baseline arms are
indistinguishable in $\rlimit$. Three factors compound: (1) $\rlimit$ is
dominated by the loop's theory-solving overhead, so branching is not the
bottleneck and $-\rlimit$ carries little RL signal; (2) there is no
conflict \emph{headroom}---even harder instances leave VSIDS with very few
conflicts; and (3) \solver{} exposes only a hard \nextsplit{}, not a soft
activity injection, so a learned \emph{full-coverage} order cannot be blended
into VSIDS. The harness asymmetry of Section~\ref{sec:omt} further weakens any
OMT claim, which we therefore report as inconclusive/negative rather than as
a controlled comparison.

% ==========================================================================
\section{Discussion}
\label{sec:discussion}
% ==========================================================================

The study yields a compact lesson: the benefit of learned Boolean branching
factors as \emph{learnable structure} $\times$ \emph{a policy that
generalizes past the dynamic baseline}. On hard SAT both factors hold; on
SMT(LIA) the first holds (learnability is restored by the theory-atom
encoding) but the second fails, because a quasi-static full-coverage order is
dominated by VSIDS's conflict-adaptive dynamics; in the OMT loop branching is
not the bottleneck at all. Two consequences follow. First, \emph{learnability
is a necessary but not sufficient} diagnostic: a decreasing imitation loss
does not imply a useful solver policy, and papers that report only the former
may overstate their case. Second, the natural next design is
\emph{selective deferral}---have the policy emit a confidence and override
only high-confidence atoms while returning the rest to VSIDS (a soft blend
into native activity)---which directly targets the SMT root cause. Further
directions include an \textsc{unsat}-capable, conflict-driven teacher (to
cover the PHP boundary), stronger theory-aware features (LP relaxation,
theory-solver feedback), and larger training budgets, all evaluated on real
solver traces by wall-clock and PAR-2.

\paragraph{Threats to validity.}
The OMT harness is asymmetric (Section~\ref{sec:omt}); SMT(LIA) variance is
large, so magnitudes require many seeds even though the direction
(learned $<$ VSIDS) is stable; and results are reported in $\confl$/$\rlimit$
rather than wall-clock, which the mechanism's callback overhead would also
affect. The measurement protocol (Definition~\ref{def:pairing}) is designed
to keep the \emph{branching} comparison itself clean.

% ==========================================================================
\section{Conclusion}
% ==========================================================================

We turned a supported solver API into a formal object---a decision oracle
over a black-box \textsc{CDCL(T)} procedure---that injects a learned
branching policy into an unmodified \solver{} while provably preserving
soundness and completeness, together with a measurement protocol that
isolates branching quality. A graph-neural policy trained by a look-ahead
teacher and REINFORCE improves on VSIDS on hard random 3-SAT but not on
SMT(LIA) theory atoms or in the OMT loop, and we traced the boundary to the
gap between a learned quasi-static order and VSIDS's conflict-adaptive
dynamics. We hope the mechanism, the isolation protocol, and the honest
boundary---learnability is not utility---are useful to others studying
learned control of production SMT/OMT solvers.

% ==========================================================================
% 注:aaai2027.sty 已内置 \bibliographystyle{aaai2027},此处无需再声明。
\bibliography{references}

\end{document}
